Counterfactual Training

### 2024-12-18 paper/presenations/slides.qmd
upped cpus

$$
\ell_{\text{clf}}(f_\theta(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x;\theta) + \lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)
$$

- $x^\prime$ are counterfactuals of $x_{s}\subseteq x$.
- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
- Counterfactuals are recycled as adversarial examples.

@@

1. Generate `nce` counterfactuals and distribute across mini-batches.
2. For each batch compute:
    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
    - Generator loss: $\ell_{\text{gen}}(x^\prime,x;\theta)$
    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)$
3. Backpropagate all losses and update parameters.

@@

>>>>

$$
\ell_{\text{clf}}(f_\theta(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}({x^\prime}_t,x_t;\theta) + \lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)
$$

- ${x^\prime}_t$ are counterfactuals of $x_{s}\subseteq x$ with target class $t$.
- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
- Counterfactuals are recycled as adversarial examples.

@@

1. Generate `nce` counterfactuals and distribute across mini-batches.
2. For each batch compute:
    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
    - Generator loss: $\ell_{\text{gen}}({x^\prime}_t,x_t;\theta)$
    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)$
3. Backpropagate all losses and update parameters.

@@

## Encoding Domain Knowledge

**Example**: Let $f_\theta(x)=\theta^{T}x$ be a linear classifier. Suppose that featue $x_1$ is immutable (e.g. 'gender' or 'faith').

$$
\begin{aligned}
\nabla_\theta \ell_{\text{gen}}(x^\prime,x;\theta) &= \nabla_\theta ( \theta^{T}x[t] - \theta^{T}x^\prime[t]) \\
\frac{\partial \ell_{\text{gen}}}{\partial\theta_1}(x^\prime,x;\theta)&=x_1 - x_1^\prime
\end{aligned}
$$

- If $$
